```{r, message = FALSE}
library(data.table)
library(ggplot2)
library(countrycode)
library(maps)
```

Test code variables.

```{r}
MATH <- "LO.PISA.MAT"
MATH_M <- "LO.PISA.MAT.MA"
MATH_F <- "LO.PISA.MAT.FE"
READ <- "LO.PISA.REA"
READ_M <- "LO.PISA.REA.MA"
READ_F <- "LO.PISA.REA.FE"
SCI <- "LO.PISA.SCI"
SCI_M <- "LO.PISA.SCI.MA"
SCI_F <- "LO.PISA.SCI.FE"
```

Read dataset, change column names, drop the column that explains the test codes.
Examine if there are countries with incomplete data in the dataset (there aren't).

```{r}
DT <- fread("pisa2015.csv", header = TRUE, na.strings = "..")
names(DT) <- c("Country", "CountryCode", "Test", "TestCode", "Score")
DT[, Test:=NULL]

has_na <- subset(DT, is.na(Score) == TRUE)
has_na <- has_na[, .N != 9, CountryCode]
has_na[V1 != FALSE]
```

Drop NA.

```{r}
DT <- na.omit(DT)
```

Add Region and EU membership columns to the dataset.

```{r, warning=FALSE}
DT$Region <- countrycode(DT$CountryCode, "iso3c", "region")
# following hack is needed because countrycode returns "EU" or NA
# instead of true or false
DT$EU <- FALSE
DT[!is.na(countrycode(CountryCode, "iso3c", "eu28"))]$EU <- TRUE
```

Examine how many countries participated in the study from each region.

```{r}
# Create new col with how many countries participated from each region
region_cnt <- DT[, .(Participating = .N/9), Region]
# region_cnt[order(-Participating)]
# Get a count of how many countries are in each region
# using codelist, a dataframe provided by the countrycode package
all_regions <- data.table(Region = na.omit(codelist$region))
all_regions_cnt <- all_regions[, .(All = .N), Region]
# Join
region_cnt <- region_cnt[all_regions_cnt, on=.(Region)]
# replace NAs with 0
region_cnt[is.na(Participating)]$Participating <- 0
region_cnt <- region_cnt[, .(ParticipatingRatio = round(Participating/All,2)), Region]
region_cnt[order(-ParticipatingRatio)]
```

Split into two data tables: one containing the average result of both genders
and one containing a Gender column (instead of having a different test code for
the results of each gender).

```{r}
mean_scores <- DT[TestCode == MATH | TestCode == READ | TestCode == SCI]

dt_gendered <- DT[TestCode != SCI & TestCode != MATH & TestCode != READ]
dt_gendered$Gender <- "Male"
dt_gendered[TestCode == MATH_F | TestCode == READ_F | TestCode == SCI_F]$Gender <- "Female"
dt_gendered[startsWith(TestCode, MATH)]$TestCode <- MATH
dt_gendered[startsWith(TestCode, READ)]$TestCode <- READ
dt_gendered[startsWith(TestCode, SCI)]$TestCode <- SCI
```

Create a new data table with the mean result of both genders at every discipline
for every country.

```{r}
one_mean <- mean_scores[, .(Score = mean(Score)), Country]

# Change some country names to match with the ones
# returned by map_data

change_name <- function(DT, old, new){
    DT[Country == old, "Country" := new]
}

updates <- list(c("Russian Federation", "Russia"),
     c("United States", "USA"),
     c("United Kingdom", "UK"),
     c("Macedonia, FYR", "North Macedonia"),
     c("Korea, Rep.", "South Korea"))

for (pair in updates){
    change_name(one_mean, pair[1], pair[2])
}

# Load map data, get a data table of all countries
# and join it with one_mean

wmap <- setDT(map_data("world", wrap = c(-180, 180)))
all_countries <- data.table(Country = unique(wmap$region))
plot_dt <- one_mean[all_countries, on = "Country"]
```

Bin the mean scores and plot them on a world map.

```{r}
discrete_values <- seq(300, 550, 50)
plot_dt$Discrete = cut(plot_dt$Score, breaks = discrete_values)

legend <- c()
for (i in 1:(length(discrete_values) - 2)){
    legend <- append(legend, paste(discrete_values[i], "-", discrete_values[i+1]))
}
legend <- append(legend, c("> 500", "No Data"))



p <- ggplot(plot_dt, aes(map_id = Country)) +
    geom_map(aes(fill = Discrete), map = wmap, color = "black", linewidth = 0.05) + expand_limits(x = wmap$long, y = wmap$lat) + 
    scale_fill_brewer(palette = "RdYlBu", na.value = "lightgrey", name = "Mean Exam Score",
                      labels = legend) + 
    coord_cartesian() + 
    theme_void() +
    theme(legend.position = "bottom")
p
# ggsave("pisa_world.svg", plot = p, width = 14, height = 7)
# ggsave("pisa_world.png", plot = p, width = 14, height = 7, bg = "white")
```
Examine the mean exam score in histogram form.

```{r}
p <- ggplot(na.omit(plot_dt)) + 
    geom_histogram(aes(x = Score, fill=Discrete), binwidth = 10, color = "black") + 
    scale_y_continuous(breaks = seq(0,20,2), labels=seq(0,20,2)) + 
    scale_fill_brewer(palette = "RdYlBu") +
    theme(legend.position = "none") + 
    labs(x = "Mean Exam Score", y = "Count")
p 
# ggsave("mean_hist.png", p, width = 7, height = 4)
```
Same plot, but with a density curve added.

```{r}
p <- ggplot(na.omit(plot_dt)) + 
    geom_histogram(aes(x = Score, y = after_stat(density)), bins = 20, color = "deepskyblue", fill = "white") + 
    geom_density(aes(x = Score), color = "mediumblue") + 
    labs(x = "Mean Exam Score", y = "Density")
p 
# ggsave("hist_density.png", p, width = 7, height = 4)
```

Boxplots of the performance of each region.

```{r}
test_labels <- c("Math", "Reading", "Science")
names(test_labels) <- c(MATH, READ, SCI)

p <- ggplot(mean_scores) + 
    geom_boxplot(aes(x = TestCode, y = Score)) + 
    facet_wrap(facets = vars(Region)) + 
    scale_x_discrete(labels = test_labels) + 
    labs(x = "Discipline", y = "Score")

p 
# ggsave("regions.png", p, width = 7, height = 4)
```

Boxplots of the performance of the EU countries compared to the rest of the world.

```{r}
# custom labels instead of true or false
facet_labels <- c(`TRUE` = "EU Members", `FALSE` = "Rest of the World")

p <- ggplot(mean_scores) + 
    geom_boxplot(aes(x = TestCode, y = Score)) + 
    # change to factor in order to fix the order of the facets
    facet_wrap(~factor(EU, levels = c(TRUE, FALSE)), labeller = as_labeller(facet_labels)) + 
    scale_x_discrete(labels = test_labels) + 
    labs(x = "Discipline", y = "Score")
p 
# ggsave("eu.png", p, width = 7, height = 4)
```
Boxplots of the performance of each gender at each discipline.

```{r}
p <- ggplot(dt_gendered) + 
    geom_boxplot(aes(x = TestCode, y = Score, color = Gender)) + 
    scale_x_discrete(labels = test_labels) + 
    labs(x = "Discipline", y = "Score")

p
# ggsave("gender-boxplots.png", p, width = 7, height = 4)
```
Same plot, but separately for each region (using ggplot's facets).

```{r}
p <- ggplot(dt_gendered) + 
    geom_boxplot(aes(x = TestCode, y = Score, color = Gender)) + 
    facet_wrap(~Region) + 
    scale_x_discrete(labels = test_labels) + 
    labs(x = "Discipline", y = "Score")

p
# ggsave("gender-facets.png", width = 7, height = 4)
```

Same plot, but seperately for EU countries and the rest of the world (again,
using ggplot's facets).

```{r}
p <- ggplot(dt_gendered) + 
    geom_boxplot(aes(x = TestCode, y = Score, color = Gender)) + 
    # change to factor in order to fix the order of the facets
    facet_wrap(~factor(EU, levels = c(TRUE, FALSE)), labeller = as_labeller(facet_labels)) + 
    scale_x_discrete(labels = test_labels) + 
    labs(x = "Discipline", y = "Score")
p 
ggsave("eu-gender-facets.png", p, width = 7, height = 4)
```


